---
title: "simcoal"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim of this code:
   Realise some simulations of a gene tree considering different coalescent model (n and $\beta$).

## ncsim -- for the n-coalescent model
### Inputs:
   'n' a positive integer for the sample size,
   
   'theta' a positive numeric for the population mutation rate,
   
   'T_total' a logical value to give the total branch length of the tree between each coalescent event (i.e. T_total will be the sum of the time given in output).

### Outputs:
   't' a vector of positive values with the time spent in each state,
   
   'Vecstates' a vector of positive integer with the transient state for this simulation,
   
   'Vecmut'a vector of positive value for the simulated SFS.

```{r}

ncsim = function(n, T_total = FALSE){
  
    if (n<=0 || length(n)!=1 || round(n)!=n){
    stop('n should be a positive integer')
  }
  
  if (!is.logical(T_total)){
    stop('T_total should be a logical value')
  }
  
  # Initialisation
  times = NULL; T_tot = NULL
  
  k = n  # number of distinct lineages
  count = 1  # because k will not goes just from 1 to size 
  Matstates = matrix(0, nrow=1, ncol=n)
  Matstates[1, ] = c(n, rep(0, n-1))
  Vecnext=Matstates[1,]
  Vecmut = rep(0, n-1)  # the number of mutation (in SFS)

  # This loop will continue until reaching the MRCA
  # (until just one lineage)
  while (k > 1){
 
    # plot(Vecrate)
    # Each coalescent time follow a exp dsitribution
    # with a parameter lambda given by the rate matrix

    lambda = k*(k-1)/2
    times[count] = rexp(n=1, rate=lambda)
    
    if(T_total == TRUE){T_tot[count] = times[count]*k} 
           
    # Need this condition because the last coalescent leads to MRCA
    # and is not represented in the state matrix
    # Vector of probability to go to each other states

    inext = NULL
    # sample two times instead of one (with size 2)
    # to remove 2 of vecnext if the sample choose 
    # two times the same lineage size
    for (i in c(1,2)){
      inext[i] = sample(x = 1:n, size = 1, prob = Vecnext, replace = T) 
      Vecnext[inext[i]]=Vecnext[inext[i]]-1
    }
    Vecnext[sum(inext)] = Vecnext[sum(inext)]+1
    
    count = count + 1
    Matstates = rbind(Matstates, Vecnext, deparse.level = 0)
    k = k - 1
  }
  # give a vector of mutation rate (corresponding to the SFS)
  
  
  
  if(T_total == T){times = T_tot}
  return(list(times, Matstates))
}
```

```{r}
mutsim = function(simtree, theta = 10){
  
  mutrate = matrix(simtree[[1]]*(simtree[[2]][-nrow(simtree[[2]]),-ncol(simtree[[2]])])*(theta/2), ncol=ncol(simtree[[2]])-1)
  mutrate = apply(mutrate, 2, sum)


  Vecmut = rpois(n = n-1, lambda = mutrate)
  return (Vecmut)
}
```
## Examples

These are a few examples of how to use this function,
example1 and example2 are exactly the same except for the output
of the time. While the first one gives the $\tau_k$, meaning the times spend with k_lineages, the second one gives the $T_{total_k}$ at each time step, i.e. the sum of all lineage branch length with k_lineages.

```{r}
# with T_MRCA

example1 = ncsim(n = 10)
print(example1)
plot(example1[[1]])
print(mutsim(example1))
``` 

```{r}
# with T_Total

example2 = ncsim(n = 10, theta = 100, T_total = T)
print(example2)
plot(example2[[1]])
print(mutsim(example2))
```

Considering many simulations, we can observe the mean SFS of the n-coalescent model. In particular this simulation confirmms that the number of singletons, doubletons, ... follows the formula presented in ,
$$
  \xi_i = \frac{\theta_i}{i}
$$

In that case if we take $n=10$, $\theta=100$ and the number of simulation is 10000, we obtain the following plot
```{r}
n=10
replicate = 1000
simulationSFS = matrix(0,ncol=n-1, nrow=replicate)
for (i in 1:replicate){
  simul = ncsim(n = n)
  simulationSFS[i,] = mutsim(simul)
}

meanSFS = apply(simulationSFS,2,mean)
q1SFS = apply(simulationSFS,2,quantile,0.05)
q2SFS = apply(simulationSFS,2,quantile,0.95)
print(meanSFS)

plot(q2SFS, type = 'l', ylab = 'mean mutation number', xlab = 'lineage size', ylim = c(0,max(q2SFS)))
lines(1:(n-1),meanSFS,col='red')
lines(q1SFS)
```

# Variation of the function for other coalescent model

## bcsim -- For the $\beta$-coalescent

### Inputs:

'n' a positive integer for the sample size,

'alpha' the alpha parameter for the beta coalescent as $\Lambda \sim \beta(\alpha,\, 2-\alpha)$ should be in the interval $[1,\,2)$

'theta' a positive numeric for the population mutation rate,
   
'T_total' a logical value to give the total branch length of the tree between each coalescent event (i.e. T_total will be the sum of the time given in output).

### Outputs:
   't' a vector of positive values with the time spent in each state,
   
   'Vecstates' a vector of positive integer with the transient state for this simulation,
   
   'Vecmut'a vector of positive value for the simulated SFS.

This function use the LambdaB() function to calculate the Lambda of the multiple-merger coalescent (according to a Beta). LambdaB() ask the value of $m$, $k$ and $\alpha$ and return the value of the $\Lambda$.

The parameter alpha goes from $[1,2)$, with $1$ is the snisn model, and as $n$ tends to $2$ we approach a Kingman's coalescent.

```{r}

LambdaB = function(m, k, alpha=1.9999){
  lmk=beta(k-alpha, m-k+alpha)/beta(alpha, 2-alpha)
  return(lmk)
  }


bcsim = function(n, alpha = 1.9999, T_total = FALSE){
  
  if (n<=0 || length(n)!=1 || round
      
      (n)!=n){
    stop('n should be a positive integer')
  }
  
  if (!is.logical(T_total)){
    stop('T_total should be a logical value')
  }
  
  if (alpha<1 || alpha >= 2 || length(alpha)!= 1){
    stop('alpha should be a positive numeric between [1, 2)')
  }
  
  times = NULL; T_tot = NULL; k = n; count = 1 
  Matstates = matrix(0, nrow=1, ncol=n)
  Matstates[1, ] = Vecnext = c(n, rep(0, n-1))
  Vecmut = rep(0, n-1)  

  while (k > 1){
    Vecrate = rep(0,n)
    for (i in 2:k){Vecrate[i]=(LambdaB(k, i, alpha = alpha))*choose(k, i)}
    lambda=sum(Vecrate)
    times[count] = rexp(n=1, rate=lambda)
    kmerger=sample(1:n, 1, prob = Vecrate/lambda)
    
    
    if(T_total == TRUE){T_tot[count] = times[count]*k} 

    inext = NULL
    for (i in 1:kmerger){
      inext[i] = sample(x = 1:n, size = 1, prob = Vecnext, replace = T) 
      Vecnext[inext[i]]=Vecnext[inext[i]]-1
    }
    Vecnext[sum(inext)] = Vecnext[sum(inext)]+1
    count = count + 1
    Matstates = rbind(Matstates, Vecnext, deparse.level = 0)
    k = k - (kmerger-1)
  }

  if(T_total == T){times = T_tot}
  return(list(times, Matstates))
}
```

## Example of the SFS for the $\beta$-coalescent
```{r}

n=100
replicate = 1000
simulationSFS = matrix(0,ncol=n-1, nrow=replicate)
for (i in 1:replicate){
  simul = bcsim(n = n, alpha = 1.2)
  simulationSFS[i,] = mutsim(simul)
}

meanSFS = apply(simulationSFS,2,mean)
meanSFS = meanSFS/sum(meanSFS)
q1SFS = apply(simulationSFS,2,quantile,0.05)
q2SFS = apply(simulationSFS,2,quantile,0.95)
q1SFS = q1SFS / sum(q1SFS)
q2SFS = q2SFS / sum(q2SFS)
print(meanSFS)

plot(q2SFS, type = 'l', ylab = 'mean mutation number', xlab = 'lineage size', ylim = c(0,max(q2SFS)))
lines(1:(n-1),meanSFS,col='red')
lines(q1SFS)

```


## Just a thought about the singleton distribution

```{r}
singletondistr = function(simultree, xton = 1, theta = 10){
  n = ncol(simultree[[2]])
  individual = rep(0, n)
  for (i in 1:length(simultree[[1]])){
    nconcerned = simultree[[2]][i, xton]
    individual[1:nconcerned] = individual[1:nconcerned]+simultree[[1]][i]
  }
  individual = individual/sum(individual)
  #individual = rpois(n,individual*theta)
  return(individual)
}
```

```{r}
n=100
replicate = 1000

neutral = matrix(0,ncol=n,nrow=replicate)
MMC = matrix(0,ncol=n, nrow=replicate)

for (i in 1:replicate){
  sim_neutral = ncsim(n = n)
  neutral[i,] = singletondistr(sim_neutral, 1, 10)
  
  sim_MMC = bcsim(n = n, alpha = 1)
  MMC[i,] = singletondistr(sim_MMC, 1, 10)
}

neutral = apply(neutral,2,mean)
neutral = neutral/sum(neutral)

MMC = apply(MMC,2,mean)
MMC = MMC/sum(MMC)

plot(neutral, type = 'l')
lines(MMC, col = 'red')
```
```{r}
n=100

neutral = matrix(0,ncol=n,nrow=replicate)
MMC = matrix(0,ncol=n, nrow=replicate)

for (i in 1:n){
  sim_neutral = bcsim(n = n, alpha=1.2)
  neutral[i,] = singletondistr(sim_neutral, i, 10)
}

require(plot3D)

persp3D(z = neutral[1:5,1:5], theta = 120)
```

```{r}
n=200
replicate = 100

neutral = matrix(0,ncol=n,nrow=replicate)
MMC = matrix(0,ncol=n, nrow=replicate)

for (i in 1:replicate){
  sim_neutral = ncsim(n = n)
  neutral[i,] = singletondistr(sim_neutral, 10)
  
  sim_MMC = bcsim(n = n, alpha = 1.5)
  MMC[i,] = singletondistr(sim_MMC, 10)
  print(i)
}

neutralx = neutral[,1]/apply(neutral,1,sum)
neutraly = apply(neutral[,5:(n)],1,sum)/apply(neutral,1,sum)

MMCx = MMC[,1]/apply(MMC,1,sum)
MMCy = apply(MMC[,5:(n)],1,sum)/apply(MMC,1,sum)

plot(neutralx, neutraly)
points(MMCx, MMCy, col = 'red')

```

## Example of the SFS for the $\beta$-coalescent
```{r}

n=100
replicate = 1000
nsimulationSFS = matrix(0,ncol=n-1, nrow=replicate)
bsimulationSFS = matrix(0,ncol=n-1, nrow=replicate)

for (i in 1:replicate){
  simul = ncsim(n = n)
  nsimulationSFS[i,] = mutsim(simul)
}
for (i in 1:replicate){
  simul = bcsim(n = n, alpha = 1.2)
  bsimulationSFS[i,] = mutsim(simul)
}

nmeanSFS = apply(nsimulationSFS,2,mean)
nmeanSFS = nmeanSFS/sum(nmeanSFS)

bmeanSFS = apply(bsimulationSFS,2,mean)
bmeanSFS = bmeanSFS/sum(bmeanSFS)


plot(bmeanSFS, type = 'l', ylab = 'mean mutation number', xlab = 'lineage size', col='red')
lines(nmeanSFS)

```



## Example of the SFS for the $\beta$-coalescent
```{r}

n=200
replicate = 1000
nsimulationSFS = matrix(0,ncol=n-1, nrow=replicate)
bsimulationSFS = matrix(0,ncol=n-1, nrow=replicate)

for (i in 1:replicate){
  simul = ncsim(n = n)
  nsimulationSFS[i,] = mutsim(simul)
}
for (i in 1:replicate){
  simul = bcsim(n = n, alpha = 1.2)
  bsimulationSFS[i,] = mutsim(simul)
}

xn = nsimulationSFS[,1]/apply(nsimulationSFS,1,sum)
yn = apply(nsimulationSFS[,100:(n-1)],1,sum)/apply(nsimulationSFS,1,sum)

xb = bsimulationSFS[,1]/apply(bsimulationSFS,1,sum)
yb = apply(bsimulationSFS[,100:(n-1)],1,sum)/apply(bsimulationSFS,1,sum)

plot(xn,yn, xlim=c(0,1), ylim=c(0,1))
points(xb,yb, col='red')
```